Parametricity for Primitive Nested Types and GADTs

### 2020-12-30 free-theorems-nesteds/20-mscs.tex
fine-tuning journal draft

\keywords{MANDATORY list of keywords}

@@

This paper is an extended version of the conference
paper~\cite{jgj21}. Compared to that paper, this one includes all
proofs, as well as more examples. The discussion of GADTs in
Section~\ref{sec:GADTs} is also entirely new.

@@

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout the model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.

@@

We have constructed a parametric model for a calculus providing
primitives for constructing nested types directly as fixpoints, rather
than representing them via their Church encodings. We have also used
the Abstraction Theorem for this model to derive free theorems for
nested types. This was not possible before~\cite{jp19} because such
types were not previously known to have well-defined interpretations
in locally finitely presentable categories (here, $\set$ and $\rel$),
and, to our knowledge, no term calculus for them existed either. The
key to obtaining our parametric model is the delicate threading of
functoriality and its accompanying naturality conditions throughout
the model construction.

We were surprised to find that, although GADTs were shown
in~\cite{jp19} to have appropriately cocontinuous functorial semantics
when interpreted using left Kan extensions, our model construction
does not extend to give a parametric model for them. It may be
possible to modify the categories of relations and relation
transformers so that these are always surjective on their domans and
codomains; this would prohibit situations like that in
Example~\ref{ex:no-rel-transf}, and might therefore allow us to
recover our IEL and, ultimately, our Abstraction Theorem for GADTs.
If this turns out to be possible, then extending this new model
construction to locally $\lambda$-presentable categories for $\lambda
> \omega$ would make it possible to handle broader classes of
GADTs. (Interestingly, $\lambda > \aleph_1$ is required to interpret
even common GADTs.) We could even attempt to carry out our
construction in locally $\lambda$-presentable cartesian closed
categories (lpcccs) $\cal C$ whose categories of (abstract) relations,
obtained by pullback as in~\cite{jac99}, are also lpcccs and are
appropriately fibred over $\cal C$. This would give a framework for
constructing parametric models for calculi with primitive GADTs that
is based on locally $\lambda$-presentable fibrations, for some
appropriate definition thereof.

The expressivity of folds for nested types has long been a vexing
issue (see, e.g.,~\cite{bm98}), and this is naturally inherited by the
calculus presented here. Since it codes all recursion using standard
folds, and since folds for nested types must return natural
transformations, many standard functions over nested types cannot be
represented in our calculus. Another important direction for future
work would therefore be to improve its expressivity by adding, say,
generalized folds~\cite{bp99}, Mendler iterators~\cite{amu05}, or
otherwise extend standard folds to express computations whose results
are not natural transformations. In particular, we may wish to add
term-level fixpoints as, e.g., in~\cite{pit00}. This would require the
categories interpreting types to be not just locally
$\lambda$-presentable, but to support some kind of domain structure as
well. At the moment it seems that such an endeavor will have to wait
for a generalization of the results presented here to at least locally
$\omega_1$-presentable categories: $\omega$-CPO, the most natural
category of domains to replace $\set$, is not locally finitely
presentable.

>>>>

\keywords{Nested types, GADTs, categorical semantics, parametricity}

@@

This paper extends the conference paper~\cite{jgj21} by including
proofs of all theorems, as well as more examples. The discussion of
GADTs in Section~\ref{sec:GADTs} is also entirely new.

@@

Note that since the body F of a type $(\mu \phi. \lambda
\ol\alpha. F)\ol{G}$ can only be functorial in $\phi$ and the
variables in $\ol{\alpha}$, the representation of
$\mathit{List}\,\alpha$ as the ADT $\mu \beta. \,\onet + \alpha \times
\beta$ cannot be functorial in $\alpha$. By contrast, if
$\mathit{List}\,\alpha$ is represented as the nested type $(\mu
\phi. \lambda \beta.\,\onet + \beta \times \phi \beta)\,\alpha$ then
we can choose $\alpha$ to be a functorial variable or not when forming
the type. This observation holds for other ADTs as well; for example,
if $\mathit{Tree}\,\alpha\,\gamma = \mu \beta. \alpha + \beta \times
\gamma \times \beta$, then $\alpha, \gamma; \emptyset \vdash
\mathit{Tree}\,\alpha\,\gamma$ is well-formed, but $\emptyset; \alpha,
\gamma \vdash \mathit{Tree}\,\alpha\,\gamma$ is not. And it also
applies to some non-ADT types, such as $\mathit{GRose}\,\phi\,\alpha =
\mu \beta. \onet + \alpha \times \phi\beta$, in which $\phi$ and
$\alpha$ must both be non-functorial variables.  It is in fact
possible to allow ``extra'' $0$-ary functorial variables in the body
of $\mu$-types (functorial variables of higher arity are the real
problem), which would allow the first-order representations of ADTs to
be functorial. However, doing this requires some changes to the
formation rule for $\mu$-types, as well as the delicate threading of
some additional conditions throughout our model construction.  But
since we can always use an ADT's (semantically equivalent)
second-order representation when functoriality is needed, disallowing
such ``extra'' variables does not negatively impact the expressivity
of our calculus. We therefore pursue the simpler syntax here.

@@

We have constructed a parametric model for a calculus providing
primitives for constructing nested types directly as fixpoints, rather
than representing them via their Church encodings. We have also used
the Abstraction Theorem for this model to derive free theorems for
nested types. This was not possible before~\cite{jp19} because such
types were not previously known to have well-defined interpretations
in locally finitely presentable categories (here, $\set$ and $\rel$),
and, to our knowledge, no term calculus for them existed either. The
key to obtaining our parametric model is the delicate threading of
functoriality and its accompanying naturality conditions throughout
our model construction.

We were surprised to find that, although GADTs were shown
in~\cite{jp19} to have appropriately cocontinuous functorial semantics
in terms of left Kan extensions, our model construction does not
extend to give a parametric model for them. It may be possible to
modify the categories of relations and relation transformers so that
these are always surjective on their domans and codomains; this would
prohibit situations like that in Example~\ref{ex:no-rel-transf}, and
might therefore allow us to recover our IEL and, ultimately, an
Abstraction Theorem appropriate to GADTs.  If this turns out to be
possible, then generalizing the resulting model construction to
locally $\lambda$-presentable categories for $\lambda > \omega$ would
make it possible to handle broader classes of GADTs. (As shown
in~\cite{jp19}, $\lambda > \omega_1$ is required to interpret even
common GADTs.) We could even attempt to carry out our construction in
locally $\lambda$-presentable cartesian closed categories (lpcccs)
$\cal C$ whose categories of (abstract) relations, obtained by
pullback as in~\cite{jac99}, are also lpcccs and are appropriately
fibred over $\cal C$. This would give a framework for constructing
parametric models for calculi with primitive GADTs that is based on
locally $\lambda$-presentable fibrations, for some appropriate
definition thereof.

The expressivity of folds for nested types has long been a vexing
issue (see, e.g.,~\cite{bm98}), and this is naturally inherited by the
calculus presented here. Since it codes all recursion using standard
folds, and since folds for nested types must return natural
transformations, many standard functions over nested types cannot be
represented in this calculus. Another important direction for future
work is therefore to improve the expressivity of our calculus by
adding, say, generalized folds~\cite{bp99}, or Mendler
iterators~\cite{amu05}, or otherwise extending standard folds to
express computations whose results are not natural transformations. In
particular, we may wish to add term-level fixpoints as, e.g.,
in~\cite{pit00}. This would require the categories interpreting types
to be not just locally $\lambda$-presentable, but also to support some
kind of domain structure. At the moment it seems that such an endeavor
will have to wait for a generalization of the results presented here
to at least locally $\omega_1$-presentable categories, however:
$\omega$-CPO, the most natural category of domains to replace $\set$,
is not locally finitely presentable.
